# Loading data

Reading in data

```{r, cache=TRUE}
co_traffic_stops<-read_csv("https://www.dropbox.com/s/r7oydximjp46ylh/co_statewide_2020_04_01.csv?dl=1")
```

Printing contents of ```co_traffic_stops``` object

```{r}
co_traffic_stops
```

# Cleaning and filtering data

## Create a "Year" field

```{r}
# Create "year" field
co_traffic_stops<-co_traffic_stops %>% mutate(Year=substr(co_traffic_stops$date, 1,4))
```

## Filter by year

```{r}
# Filter 2010 observations and assign to a new object
co_traffic_stops_2010<-co_traffic_stops %>% filter(Year==2010)
```

## View the newly created ```co_traffic_stops_2010``` object 

```{r}
# Print contents of "co_traffic_stops_2010" object
co_traffic_stops_2010
```

# Transforming Data

## Tabulate county-level count of traffice stops by race

```{r}
# Compute county-level count of traffic stops by race
co_county_summary<-co_traffic_stops_2010 %>% 
                    group_by(county_name) %>% 
                    count(subject_race) 
```

## Reshape the data

```{r}
co_county_summary_wide<-co_county_summary %>% 
                        pivot_wider(names_from=subject_race, values_from=n)
```

## Calculate total stops for each county in ```co_county_summary_wide```

```{r}
co_county_summary_wide<-co_county_summary_wide %>% 
                        rowwise() %>% 
                        mutate(total_stops=sum(c_across(where(is.integer)), na.rm=TRUE)) 
```

```{r}
co_county_summary_wide
```

## Clean ```co_county_summary_wide``` and assign to new object

```{r}
co_county_black_stops<-co_county_summary_wide %>%
                        select(county_name, black, total_stops) 
                        
```


```{r}
co_county_black_stops<-co_county_black_stops %>%
                        rename(black_stops=black)
```


```{r}
co_county_black_stops<-co_county_black_stops %>% 
                        filter(county_name!="NA")

```


```{r}
co_county_black_stops
```

# Defining an index of racial bias in traffic stops

## Read in and join 2010 census data to ```co_county_black_stops```

Read in census data

```{r}
# Read in census data and assigns to object named "co_counties_census_2010"
co_counties_census_2010<-read_csv("https://www.dropbox.com/s/kq94707fmcdbc0w/co_county_decennial_census.csv?dl=1") 
```

```{r}
# Prints contents of "co_counties_census_2010"
co_counties_census_2010
```

## Join census data to ```co_counties_census_2010```

```{r}
co_counties_census_trafficstops<-full_join(co_county_black_stops, co_counties_census_2010,
                                           by=c("county_name"="County"))
```

```{r}
co_counties_census_trafficstops
```

## Define the variables that will be used in the bias index

```{r}
co_counties_census_trafficstops<-co_counties_census_trafficstops %>% 
                                    mutate(black_stop_pct=((black_stops/total_stops)*100),
                                           black_pop_pct=((total_black_pop_over17/total_pop_over17)*100))
```

```{r}
co_counties_census_trafficstops
```

## Calculate the bias index

```{r}
co_counties_census_trafficstops<-co_counties_census_trafficstops %>% 
                                  mutate(excess_stops_index=black_stop_pct-black_pop_pct)
```

```{r}
co_counties_census_trafficstops
```

## Compute summary statistics for the bias index

```{r}
describe(co_counties_census_trafficstops$excess_stops_index)
```

## Visualize county-level variation in the bias index using ```ggplot```

```{r}
co_counties_census_trafficstops<-co_counties_census_trafficstops %>% 
                                 mutate(County=str_remove(county_name, " County"))
```

```{r}
bias_graph<-co_counties_census_trafficstops %>% 
            drop_na(excess_stops_index) %>% 
            ggplot()+
            geom_col(aes(x=reorder(County, excess_stops_index), y=excess_stops_index))+
            coord_flip()+
            labs(title="Racial Discrimination in Police Stops, by County (Colorado)", x="County", y="Bias Index=\nBlack Percentage of Overall Traffice Stops – Black Percentage of Overall Over-17 Population")+
            theme(axis.title.x = element_text(size = 9))+      
            scale_y_continuous(breaks=c(-10, -7.5, -5, -2.5, 0, 2.5, 5, 7.5))+
            expand_limits(y=c(-10,10))+
            theme(plot.title=element_text(hjust=0.5)) 
```

```{r, fig.width=8, fig.height=10}
bias_graph
```

# Mapping the bias index

## Read in and view the shapefile of CO counties

```{r, echo=-1}
setwd("/Users/adra7980/Documents/CU_workshops/gis/data/tl_2019_08_county")
# Reads in shapefile and assigns to object named "co_counties_shapefile"
co_counties_shapefile<-st_read("tl_2019_08_county.shp")
```

```{r}
co_counties_shapefile
```

```{r, fig.asp=0.5, echo=-1}
tmap_mode("plot")
tm_shape(co_counties_shapefile)+
  tm_polygons()
```

```{r}
co_counties_map<-tm_shape(co_counties_shapefile)+
                    tm_polygons()
```

```{r, fig.asp=0.5}
co_counties_map
```

```{r}
tmap_mode("view")
```

```{r}
co_counties_map
```

```{r}
tmap_mode("plot")
```

```{r, fig.asp=0.5}
co_counties_map
```

## Join ```co_counties_census_trafficestops``` to the shapefile of Colorado counties

```{r}
county_shapefile_biasIndex<-full_join(co_counties_shapefile, co_counties_census_trafficstops, 
                                   by="GEOID")
```

```{r}
county_shapefile_biasIndex
```

## Display the bias index on a map of Colorado counties

### Make a rough draft of a map 

```{r, fig.asp=0.5}
traffic_map_race<-tm_shape(county_shapefile_biasIndex)+
                  tm_polygons(col="excess_stops_index", 
                              palette="YlOrRd", 
                              textNA="No Data", 
                              n=5, 
                              style="jenks", 
                              midpoint=F)+
                  tm_layout(frame=FALSE, 
                            legend.outside=TRUE)

traffic_map_race
```

### Make a map with custom breaks

```{r, fig.asp=0.5}
traffic_map_race<-tm_shape(county_shapefile_biasIndex)+
                  tm_polygons(col="excess_stops_index", 
                              palette="YlOrRd", 
                              textNA="No Data", 
                              n=5, 
                              breaks=c(-10,-5, 0, 2, 4, 5),
                              midpoint=F)+
                  tm_layout(frame=FALSE, 
                            legend.outside=TRUE)
traffic_map_race
```

### Make a map with custom colors 

```{r, fig.asp=0.5}
my_colors<-c("white", "peachpuff", "red1", "red4", "navy")
traffic_map_race<-tm_shape(county_shapefile_biasIndex)+
                  tm_polygons(col="excess_stops_index", 
                              palette=my_colors, 
                              textNA="No Data", 
                              n=5, 
                              breaks=c(-10,-5, 0, 2, 4, 5))+
                  tm_layout(frame=FALSE, 
                            legend.outside=TRUE)
traffic_map_race
```

### Make a categorical map 

Define categorical column 

```{r}
county_shapefile_biasIndex<-
  county_shapefile_biasIndex %>% 
            mutate(apparent_bias=ifelse(excess_stops_index>0, "Apparent Bias", "No Apparent Bias"))
```

```{r, fig.asp=0.5}
categorical_map<-tm_shape(county_shapefile_biasIndex)+
                    tm_polygons(col="apparent_bias", title="", pal=c("orangered1", "white"), 
                                textNA="No Data")+
                    tm_layout(frame=FALSE, 
                              legend.outside=TRUE)
                    
categorical_map
```


# Refining and formatting the maps of the Colorado traffic-stop bias index

## Refining the map of the continuous bias variable

```{r, echo=F}
county_shapefile_biasIndex<-county_shapefile_biasIndex %>% relocate(NAME)
```

```{r, fig.asp=0.5}
my_colors<-c("white", "peachpuff", "red1", "red4", "navy")
traffic_map_race<-tm_shape(county_shapefile_biasIndex)+
  tm_polygons(col="excess_stops_index", 
              palette=my_colors,
              title="(Black % of Traffic Stops) - (Black % of Population)",
              textNA="No Data", 
              n=5, 
              breaks=c(-10,-5, 0.01, 2, 4, 5))+
  tm_layout(frame=FALSE, 
            legend.outside=TRUE,
            legend.text.size=0.68,
            legend.title.size=0.75,
            title="Traffic Stop Racial Bias Index",
            title.size=0.75,
            title.fontface = 2,
            main.title="Racial Bias in Traffic Stops, by County (Colorado)",
            main.title.position=0.03,
            main.title.size=1,
            attr.outside=TRUE)+
  tm_credits("Map Author: NAME\nData Sources: 2010 Decennial Census via tidyverse,\nStanford Open Policing Project,\nColorado GeoLibrary ", # Sets text for map credits
                    position=c(0.02,0.01), # Specifies location of map credits
                    size=0.38)

traffic_map_race
```


```{r}
traffic_map_race_labeled<-traffic_map_race+
                          tm_text("NAME", size=0.30, fontface=2)

traffic_map_race_labeled
```

```{r}
tmap_mode("view")

traffic_map_race
```

```{r}
tmap_mode("plot")
```

## Refining the map of the categorical bias variable 

```{r}
categorical_map<-tm_shape(county_shapefile_biasIndex)+
                    tm_polygons(col="apparent_bias", title="", pal=c("orangered1", "white"), 
                                textNA="No Data")+
                    tm_layout(frame=FALSE, 
                              legend.outside=TRUE,
                              main.title="Racial Bias in Traffic Stops, by County (Colorado)",
                              main.title.position=0.03,
                              main.title.size=1,
                              attr.outside=TRUE)+
                 tm_credits("Map Author: NAME\nData Sources: 2010 Decennial Census via tidyverse,\nStanford Open Policing Project,\nColorado GeoLibrary ", # Sets text for map credits
                    position=c(0.02,0.01), # Specifies location of map credits
                    size=0.38)+
  tm_text("NAME", size=0.30, fontface=2)
      
                    
categorical_map
```

# Exporting maps

# Summary script


lower bound on legend is inclusive, upper bound exclusive




[https://geo.colorado.edu/catalog/47540-5e712aeda3d91e0009f59fc7](https://geo.colorado.edu/catalog/47540-5e712aeda3d91e0009f59fc7)



